---
title: "Metabolite enrichment"
author: "Vincent ROCHER & Juliette COOKE"
format: revealjs
editor: visual
---

```{r}
#| label: load-packages
#| include: false
#| 
library(clusterProfiler)
library(tidyverse)
library(FELLA)
```

# Overview

## Introduction

-   Pathway enrichment methods were originally developed for gene enrichment

-   Are also widely used with metabolites and metabolomics data (with caution!)

## Differences vs genes

-   Lower metabolome coverage vs genome coverage, less data in general

-   No Gene Ontology equivalent, ChEBI ontology is close =\> gaining functional information is more difficult

-   Metabolite identifier mapping is a major issue: metabolites have multiple names, IDs, isomers

-   Bias depending on input data used, more so for metabolomics

## Methods

-   ORA: Over-representation analysis (Fisher's exact test)

-   MSEA: Metabolite set enrichment analysis, often referred to as GSEA even for metabolites = a Functional Class Scoring (FCS) method

-   Topology-based methods: uses networks of pathways in addition to other methods

## Tools and Databases

### Databases and IDs

-   KEGG, PubChem, ChEBI, MeSH, HMDB are databases which all have their own metabolite IDs, some with links to each other

-   Converting IDs: MetaboAnalyst can convert from names/IDs to many different IDs

### Tools

-   R: `clusterprofiler`, `FELLA`

-   In addition to R, Python can also be used for enrichment analyses: Python packages include `gseapy` with `gsea()` or `prerank()` functions

-   MetaboAnalyst: for online analyses

-   Metexplore for visualisation using metabolic networks

### Pathway sets

-   KEGG, Reactome, BioCyc, BiGG/MetabolicAtlas

-   Pathway definitions are relatively arbitrary and depend on the database

# Enrichment methods

## Over-representation analysis (ORA)

-   Uses Fisher's exact test to test for overlap in sets

-   3 inputs:

    -   a collection of pathways

    -   a list of metabolites of interest (i.e. filtered on p-values)

    -   a background set:

        -   targeted metabolomics: all assayed metabolites

        -   untargeted metabolomics: all features than can be annotated

## ORA: extra info

-   Background set is very important! Use a specific background set: realistically observable, not all known metabolites

-   Multiple-testing corrections + filtering are necessary to produce list of metabolites of interest

## Metabolite set enrichment analysis (MSEA/GSEA)

-   Ranks all metabolite values and tests if closely ranked metabolites enrich the same pathway

-   Provides an Enrichment Score of each pathway which indicates up or down regulation

-   Doesn't require a threshold on metabolite values as it uses the entire list of metabolites of interest

-   Doesn't need a background set (background set = all input metabolites)

## GSEA: extra info

-   Provides a list of leading metabolites which help enrich each pathway (where in clusterprofiler?)

-   Still need to filter on p-value for significantly enriched pathways

-   Depending on input values/scores, interpretation can be different!

## General warnings

-   Sensitive to metabolite misidentification

-   Use organism-specific pathway set if available

-   If possible, run multiple times with different pathway databases and form a consensus

# Enrichment tools

## Clusterprofiler

-   ORA?

-   GSEA/MSEA

```{r}
#cond = zscores %>% as.data.frame() %>% select(2)  %>% rownames_to_column("id")
#write.table(cond %>% rename("score" = 2), paste0("data/example_metab_data_",names(cond)[2],".tsv"), sep="\t", quote = F, row.names = F)
```

### GSEA example: input data

::: columns
::: {.column width="45%"}
Pathway sets

```{r}
pathway_sets = read.table("/home/juliette/these/data/simulated_PA/dict/metabolite_pathways.tsv", sep="\t", header = T) %>% 
  mutate(metabolite = str_remove(metabolite, "[a-z]$")) %>% relocate(subsystem)

metabolites = read.table("data/example_metab_data_Acylglycerides_metabolism.tsv", sep="\t", header = T) %>% arrange(desc(score)) %>% mutate(id = as.character(id)) %>% pull(score,id) %>% na.omit()

gsea.metab = GSEA(metabolites, TERM2GENE = pathway_sets,minGSSize = 3, pvalueCutoff = 1)
pathway_sets %>% slice(sample(1:n())) %>% head(15) %>% tibble()
```
:::

::: {.column width="10%"}
:::

::: {.column width="45%"}
Metabolomics data

```{r}
metabolites %>% enframe() %>% tibble() %>% head(15)
```
:::
:::

### GSEA example: results

```{r}
tibble((gsea.metab@result)) %>% filter(pvalue < 0.05) %>% select(ID, setSize, NES, pvalue) %>% head(8)
```

### GSEA example: plot

```{r}
enrichplot::gseaplot2(gsea.metab, geneSetID = 8, title = gsea.metab$Description[8])
```

## MetaboAnalyst

-   ORA

-   MSEA

-   Pathway topology (PT)

## FELLA

-   Uses KEGG network and a list of KEGG metabolites

-   Represents KEGG as a knowledge graph with hierarchical connections from metabolties to pathways through reactions, enzymes...

## FELLA: example

```{r}
data("FELLA.sample")
input.sample = read.table("data/epithelial_input.csv", sep= ",", header = T)
```

```{r}
#| eval: false
graph <- buildGraphFromKEGGREST(
 organism = "hsa",
 filter.path = c("01100", "01200", "01210", "01212", "01230"))

buildDataFromGraph(
 keggdata.graph = graph,
 databaseDir = "kegg_db/",
 internalDir = FALSE,
 matrices = "diffusion",
 normality = "diffusion",
 niter = 50)
```

```{r}
fella.data <- loadKEGGdata(
 databaseDir = "data/kegg_db",
 internalDir = FALSE,
 loadMatrix = NULL
 )
```

```{r}
myAnalysis = enrich(compounds = input.sample%>% pull(KEGG.id), method = "diffusion", 
    approx = "normality",
    data = fella.data)
```

```{r}
g <- generateResultsGraph(object = myAnalysis, data = fella.data)
# plot(
#     x = myAnalysis, 
#     method = "diffusion", 
#     threshold = 0.05,
#     nlimit = 150,
#     data = fella.data,
#     vertex.label.cex = 0.5)
v.size <- c(
  "1" = 10,
  "2" = 6,
  "3" = 5,
  "4" = 4,
  "5" = 2
)
v.color <- stats::setNames(grDevices::palette()[1:5], 1:5)

```

```{r fig.width=6, fig.height=7}
#| fig-width: 17
#| fig-height: 10
plotGraph(g, 
     vertex.size = v.size[V(g)$com], 
     vertex.color = v.color[V(g)$com], 
     # only show level 1 (pathway) and 5 (compound)
     vertex.label = ifelse(V(g)$com %in% c(1, 5), V(g)$label, ""), 
     vertex.label.dist = 1, 
     vertex.label.color = ifelse(V(g)$name %in% input.sample, 
                                 "indianred1", "gray20"), 
     edge.width = 3,
     nlimit = 150)
```

## Metexplore

-   Pathway analysis

-   Visualisation

# Conclusion

## References

-   Wieder C, Frainay C, Poupin N, Rodríguez-Mier P, Vinson F, Cooke J, et al. (2021) **Pathway analysis in metabolomics: Recommendations for the use of over-representation analysis.** PLoS Comput Biol 17(9): e1009105. https://doi.org/10.1371/journal.pcbi.1009105
